---
title: "Generalized Additive Models (GAMs)"
author: "AliMirzaei"
date: "11/4/2022"
output: html_document
# font-import: http://fonts.googleapis.com/css?family=Roboto|Open+Sans|Stalemate|Alex+Brush
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
Sys.setlocale(category = "LC_ALL", locale = "persian")
```

In LM we model the mean of data as a sum of linear terms:

$$y_i = \beta_0 +\sum_j \color{red}{ \beta_j x_{ji}} +\epsilon_i$$
A GAM is a sum of _smooth functions_ or _smooths_

$$y_i = \beta_0 + \sum_j \color{red}{s_j(x_{ji})} + \epsilon_i$$
where $\epsilon_i \sim N(0, \sigma^2)$, $y_i \sim \text{Normal}$

## Fitting a GAM in R
```r
model <- gam(y ~ s(x1) + s(x2) + te(x3, x4), # formuala describing model
             data = my_data_frame,           # your data
             method = 'REML',                # or 'ML'
             family = gaussian)              # or something more exotic
```

`s()` terms are smooths of one or more variables

`te()` terms are the smooth equivalent of *main effects + interactions*

## Smooth interactions

Two ways to fit smooth interactions

1. Bivariate (or higher order) thin plate splines
    * `s(x, z, bs = 'tp')`
    * Isotropic; single smoothness parameter for the smooth
	* Sensitive to scales of `x` and `z`
2. Tensor product smooths
    * Separate marginal basis for each smooth, separate smoothness parameters
	* Invariant to scales of `x` and `z`
	* Use for interactions when variables are in different units
	* `te(x, z)`

## Tensor product smooths

There are multiple ways to build tensor products in *mgcv*

1. `te(x, z)`
2. `t2(x, z)`
3. `s(x) + s(z) + ti(x, z)`

`te()` is the most general form but not usable in `gamm4::gamm4()` or *brms*

`t2()` is an alternative implementation that does work in `gamm4::gamm4()` or *brms*

`ti()` fits pure smooth interactions; where the main effects of `x` and `z` have been removed from the basis

## Type of smooths

The type of smoother is controlled by the `bs` argument (think *basis*)

The default is a low-rank thin plate spline `bs = 'tp'`

Many others available



* Cubic splines `bs = 'cr'`
* P splines `bs = 'ps'`
* Cyclic splines `bs = 'cc'` or `bs = 'cp'`
* Adaptive splines `bs = 'ad'`
* Random effect `bs = 're'`
* Factor smooths `bs = 'fs'`
* Duchon splines `bs = 'ds'`
* Spline on the sphere `bs = 'sos'`
* MRFs `bs = 'mrf'`
* Soap-film smooth `bs = 'so'`
* Gaussian process `bs = 'gp'`

## Factor smooth interactions

Two ways for factor smooth interactions

1. `by` variable smooths
    * entirely separate smooth function for each level of the factor
	* each has it's own smoothness parameter
	* centred (no group means) so include factor as a fixed effect
	* `y ~ f + s(x, by = f)`
2. `bs = 'fs'` basis
    * smooth function for each level of the function
	* share a common smoothness parameter
	* fully penalized; include group means
	* closer to random effects
	* `y ~ s(x, f, bs = 'fs')`


## A bestiary of conditional distributions

A GAM is just a fancy GLM

Simon Wood & colleagues (2016) have extended the *mgcv* methods to some non-exponential family distributions


* `binomial()`
* `poisson()`
* `quasipoisson()`
* `Gamma()`
* `inverse.gaussian()`
* `nb()`
* `tw()`
* `mvn()`
* `multinom()`
* `betar()`
* `scat()`
* `gaulss()`
* `ziplss()`
* `twlss()`
* `cox.ph()`
* `gamals()`
* `ocat()`

## Libraries

```{r libraries, message=FALSE, warning=FALSE}
if(!require('dplyr')) install.packages('dplyr')
if(!require('corrplot')) install.packages('corrplot')
if(!require('mgcv')) install.packages('mgcv')
if(!require('plotly')) install.packages('plotly')
if(!require('GGally')) install.packages('GGally')
if(!require('gratia')) install.packages('gratia')
if(!require('ggeffects')) install.packages('ggeffects')
if(!require('scico')) install.packages('scico')
if(!require('beepr')) install.packages('beepr')
if(!require('purrr')) install.packages('purrr')
if(!require('tibble')) install.packages('tibble')
if(!require('visibly')) devtools::install_github('m-clark/visibly', upgrade = "never")
if(!require('tidyext')) devtools::install_github('m-clark/tidyext', upgrade = "never")
```

## Dataset


```{r dataset, echo=FALSE, warning=FALSE}
# df <- read.csv(paste0(here::here(),'/esf98_tag_viloation(1).csv'), encoding="UTF-8")

df <- read.csv('C:/Users/Traffic/Desktop/SOHBATZADEH/esf98_tag_viloation (2).csv', encoding="UTF-8")

df <- df %>% mutate_if(is.character, as.factor) %>% 
  relocate(accident)

df$high_holiday <- ifelse(df$TAG ==  "تعطيلي بلند مدت", 1 , 0)
df$low_holiday <- ifelse(df$TAG ==  "تعطيلي کوتاه مدت", 1 , 0)
df$holiday <- ifelse(df$TAG ==  "تعطيل", 1 , 0)
df$normal <- ifelse(df$TAG ==  "عادي", 1 , 0)

cols <- c('IS_HOLIDAY'
           , 'high_holiday', 'low_holiday', 'holiday', 'normal'
          )
df[cols] <- lapply(df[cols], factor)
# df <- df %>% mutate(c.day = 1:nrow(df))
```
```{r data}
glimpse(df)
```

## Exploratory Data Analysis
```{r EDA, message=FALSE, warning=FALSE}
# knitr::kable(df %>% tidyext::num_by(vars(-JDATE,-JYEAR, -TAG, -JMONTH, -JDAY, -JDATE, -IS_HOLIDAY, - DESCRIPTION, -h_holiday, -l_holiday)), align = 'c')

knitr::kable(df %>% tidyext::cat_by(main_var = TAG, digits = 2), align = 'c')
knitr::kable(df %>% tidyext::cat_by(main_var = IS_HOLIDAY, digits = 2), align = 'c')

smooth <- function(data, mapping, ptcol, ptalpha=1, ptsize=1, linecol, ...) {
  p <- ggplot(data = data, mapping = mapping) +
    geom_point(color=ptcol, alpha=ptalpha, size=ptsize) +
    geom_smooth(color=linecol, ...)
  p
}

# ggpairs(df %>% dplyr::select(-JDATE,-JYEAR, -TAG,
#                       -JMONTH, -JDAY, -JDATE,
#                       -IS_HOLIDAY, -DESCRIPTION) %>% 
#           rename(DIST_VIOL = DISTANCE_VIOLATIONS,
#                  HVY_VEH = HEAVY_VEHICLES,
#                  SPD_VIOL = SPEED_VIOLATIONS,
#                  TTL_VEH = TOTAL_VEHICLES),
#   lower = list(
#     continuous = GGally::wrap(
#       smooth,
#       ptalpha = .25,
#       ptcol = '#D55E00',
#       ptsize = 1,
#       linecol = '#03b3ff',
#       method = 'loess',
#       se = F,
#       lwd = .5
#     )
#   ),
#   diag = list(continuous = GGally::wrap(
#     'densityDiag', color = 'gray50', lwd = .25
#   )),
#   upper = list(continuous = wrap("cor", size = 2.8)),
#   axisLabels = "none") + 
#   theme(strip.text.x = element_text(angle = 90, hjust = 0),
#            strip.text.y = element_text(angle = 0, hjust = 0),
#         strip.background.x = element_blank(),
#         strip.background.y = element_blank())

corrplot(
  cor(df %>% select(-JDATE,-JYEAR, -TAG,
                    -JMONTH, -JDAY, -JDATE,
                    -IS_HOLIDAY, -DESCRIPTION,
                    -h_holiday, -l_holiday)),
  method = 'square',
  type = 'upper',
  tl.col = 'black',
  tl.cex = 0.75, tl.srt = 90,
  col = colorRampPalette(c('purple', 'dark green'))(200)
)
```

<!-- # Model 1 -->

<!-- ```{r smooth_parameter, fig.width=10, warning=FALSE} -->
<!-- fits = purrr::map_df( -->
<!--   c(0.00001, 0.01, 1, 100, 1000000), -->
<!--   function(p) -->
<!--     tibble(sp = p, -->
<!--            JDATE = df$JDATE, -->
<!--            fits = fitted( -->
<!--              gam(accident  ~ s(SPEED_VIOLATIONS, k = 30) + -->
<!--                    s(DISTANCE_VIOLATIONS)  -->
<!--                  ,data = df, family = 'poisson', sp = p) -->
<!--            )) -->
<!-- ) %>% -->
<!--   mutate(sp = factor(sp, labels = c(0.00001, 0.01, 1, 100, 1000000),)) -->



<!-- plot_ly(data = df) %>% -->
<!--   add_markers( x = ~JDATE, -->
<!--                y = ~accident , -->
<!--                marker = list(color = '#D55E00', opacity = .2), -->
<!--                showlegend = F) %>% -->
<!--   add_lines( x = ~JDATE, y = ~ fits, color =  ~ sp, data = fits)%>% -->
<!--   theme_plotly() %>% config(displayModeBar = T) -->
<!-- ``` -->

```{r 3dscatter}

```

# Modele 1

```{r Modeling, message=FALSE, warning=FALSE}
system.time({mod_gam1 = gam(accident  ~
                             s(SPEED_VIOLATIONS, k = 12, bs = 'tp') +
                             s(DISTANCE_VIOLATIONS, k = 12, bs = 'tp') +
                             # s(SPEED_VIOLATIONS, DISTANCE_VIOLATIONS,
                             #   k = 12, bs = 'tp') +
                             # ti(police_enforcment, holiday, bs = 'fs')+
                             # s(police_enforcment, k = 12) +
                             s(JMONTH, bs = 'fs', k = 12) +
                             #ti(SPEED_VIOLATIONS, JMONTH, bs = 'fs', k = 12) +
                             # s(TAG, bs = 'fs', k = 12) +
                             # s(class4, bs = 'cr') +
                             # s(avg_class1_speed, k = 12) +
                             # s(avg_class3_speed, k = 12) +
                              ti(class4, IS_HOLIDAY, bs = 'fs') +
                             ti(police_enforcment, IS_HOLIDAY, bs = 'fs') +
                             ti(police_enforcment, JMONTH, bs = 'fs')
                              #ti(SPEED_VIOLATIONS, IS_HOLIDAY, bs = 'fs')
                           ,data = df, select = T, family = 'poisson'
                           # ,method = 'REML'
                           )
# beepr::beep(2)
})
summary(mod_gam1)
# plot_gam_check(mod_gam, scatter = TRUE)
```
# Modele 2

```{r Modeling, message=FALSE, warning=FALSE}
system.time({mod_gam2 = gam(accident  ~
                             s(SPEED_VIOLATIONS, k = 12, bs = 'tp') +
                             s(DISTANCE_VIOLATIONS, k = 12, bs = 'tp') +
                             # s(SPEED_VIOLATIONS, DISTANCE_VIOLATIONS,
                             #   k = 12, bs = 'tp') +
                             # ti(police_enforcment, holiday, bs = 'fs')+
                             # s(police_enforcment, k = 12) +
                             s(JMONTH, bs = 'fs', k = 12) +
                             #ti(SPEED_VIOLATIONS, JMONTH, bs = 'fs', k = 12) +
                             # s(TAG, bs = 'fs', k = 12) +
                             # s(class4, bs = 'cr') +
                             # s(avg_class1_speed, k = 12) +
                             # s(avg_class3_speed, k = 12) +
                              ti(class4, IS_HOLIDAY, bs = 'fs') +
                             ti(police_enforcment, IS_HOLIDAY, bs = 'fs') +
                             ti(police_enforcment, JMONTH, bs = 'fs') +
                             ti(class2_speed, IS_HOLIDAY, bs = 'fs')
                           ,data = df, select = T, family = 'poisson'
                           # ,method = 'REML'
                           )
# beepr::beep(2)
})
summary(mod_gam2)
# plot_gam_check(mod_gam, scatter = TRUE)
```

```{r aic}
AIC(mod_gam1, mod_gam2)
```

```{r acf}
res <- pred$accident - pred$value...34
forecast::Acf(res)
```
```{r fitted plot, fig.width=9}
pred1 <- as_tibble(predict(mod_gam1, newdata = df, se.fit = F,
                          type = 'response'))
pred2 <- as_tibble(predict(mod_gam2, newdata = df, se.fit = F,
                          type = 'response'))

pred <- bind_cols(df, pred1, pred2)
# %>%
#   mutate(upr = as.numeric(fit + 1.96*se.fit), lwr = as.numeric(fit - 1.96*se.fit))

plot_ly(data = pred, x = ~JDATE,
        y = ~accident, type = 'scatter', mode = 'lines',
        hoverinfo = 'text',
        hovertext = paste(
          "Date: ", pred$JDATE,
          "<br>Accident: ", pred$accident,
          "<br>GAM1: ", round(pred$value...33, digits = 0),
          "<br>GAM2: ", round(pred$value...34, digits = 0),
          "<br>Distance violations: ", pred$DISTANCE_VIOLATIONS,
          "<br>Police violations: ", pred$SPEED_VIOLATIONS,
          "<br>Police enforcment: ", pred$police_enforcment,
          "<br>Is Holiday: ", pred$IS_HOLIDAY,
          "<br>TAG: ", pred$TAG
        ),
        name = 'Data',
        yaxis = 'y') %>% 
  add_lines(x = ~JDATE,
            y = ~value...33, name='GAM1', 
            line=list(color="red", width=2)) %>%
  add_lines(x = ~JDATE,
            y = ~value...34, name='GAM2', 
            line=list(color="yellow", width=2)) %>% 
  add_trace(data = pred, x = ~JDATE,
        y = ~SPEED_VIOLATIONS, type = 'scatter', mode = 'lines',
        hoverinfo = 'text',
        hovertext = paste(
          "Date: ", pred$JDATE,
          "<br>Accident: ", pred$accident,
          "<br>GAM1: ", round(pred$value...33, digits = 0),
          "<br>GAM2: ", round(pred$value...34, digits = 0),
          "<br>Distance violations: ", pred$DISTANCE_VIOLATIONS,
          "<br>TOTAL VEHICLES: ", pred$TOTAL_VEHICLES,
          "<br>Police violations: ", pred$SPEED_VIOLATIONS,
          "<br>Police enforcment: ", pred$police_enforcment,
          "<br>Is Holiday: ", pred$IS_HOLIDAY,
          "<br>TAG: ", pred$TAG
        ),
        name = 'speed', yaxis = 'y2') %>% 
  layout(yaxis2 = list(overlaying = "y", side = "right",title="\u0641\u0648\u062a\u06cc"))
  # add_ribbons(x = ~JDATE, ymin = ~lwr, ymax = ~upr, 
  #             name='GAM 95% CI',
  #             line=list(color="#366092", opacity=0.4,
  #                       width=0)) %>% 
  # layout(legend = list(orientation = 'h', y = -0.25))

# plot_ly(data = pred, x = ~JDATE,
#         y = ~SPEED_VIOLATIONS, type = 'scatter', mode = 'lines',
#         hoverinfo = 'text',
#         hovertext = paste(
#           "Date: ", pred$JDATE,
#           "<br>Accident: ", pred$accident,
#           "<br>GAM1: ", round(pred$value...33, digits = 0),
#           "<br>GAM2: ", round(pred$value...34, digits = 0),
#           "<br>Distance violations: ", pred$DISTANCE_VIOLATIONS,
#           "<br>TOTAL VEHICLES: ", pred$TOTAL_VEHICLES,
#           "<br>Police violations: ", pred$SPEED_VIOLATIONS,
#           "<br>Police enforcment: ", pred$police_enforcment,
#           "<br>Is Holiday: ", pred$IS_HOLIDAY,
#           "<br>TAG: ", pred$TAG
#         ),
#         name = 'Data')


```

<!-- # ```{r plot, fig.width=9} -->
<!-- # plot_ly(data = df, x = ~SPEED_VIOLATIONS, -->
<!-- #         y = ~DISTANCE_VIOLATIONS, z =~accident)  -->
<!-- # ``` -->

<!-- # Modele 2 -->
<!-- ```{r Modeling2, message=FALSE, warning=FALSE} -->
<!-- system.time({mod_gam2 = gam(accident  ~ s(SPEED_VIOLATIONS, k = 12) +  -->
<!--                              s(DISTANCE_VIOLATIONS, k = 12) + -->
<!--                              # s(police_enforcment, k = 12) + -->
<!--                              s(JMONTH) + -->
<!--                              s(avg_class2_speed, k = 12) + -->
<!--                              s(avg_class3_speed, k = 12) + -->
<!--                              ti(class4, IS_HOLIDAY, bs = 'fs') + -->
<!--                              ti(HEAVY_VEHICLES, IS_HOLIDAY, bs = 'fs') + -->
<!--                              # ti(police_enforcment, IS_HOLIDAY, bs = 'fs') + -->
<!--                              ti(police_enforcment, JMONTH, bs = 'fs') -->
<!--                            ,data = df, select = T, family = 'poisson') -->
<!-- beepr::beep(2) -->
<!-- }) -->
<!-- summary(mod_gam2) -->
<!-- # plot_gam_check(mod_gam2, scatter = TRUE) -->
<!-- ``` -->
<!-- # Modele 3 -->
<!-- ```{r Modeling3, message=FALSE, warning=FALSE} -->
<!-- system.time({mod_gam3 = gam(accident  ~ -->
<!--                               s(class1_speed, k = 12) +  -->
<!--                              s(DISTANCE_VIOLATIONS, k = 12) + -->
<!--                              # s(police_enforcment, k = 12) + -->
<!--                              s(JMONTH) + -->
<!--                              s(avg_class2_speed, k = 12) + -->
<!--                              s(avg_class3_speed, k = 12) + -->
<!--                              # ti(class4, TAG, bs = 'fs') + -->
<!--                              ti(class2_speed, class2, k = 12) + -->
<!--                              s(DISTANCE_VIOLATIONS, class1, k = 12) + -->
<!--                              # ti(HEAVY_VEHICLES, IS_HOLIDAY, bs = 'fs') + -->
<!--                              s(police_enforcment, IS_HOLIDAY, bs = 'fs') + -->
<!--                              ti(police_enforcment, JMONTH, bs = 'fs') -->
<!--                            ,data = df, select = T, family = 'poisson') -->
<!-- beepr::beep(2) -->
<!-- }) -->
<!-- summary(mod_gam3) -->
<!-- # plot_gam_check(mod_gam3, scatter = TRUE) -->
<!-- ``` -->
<!-- # Modele 4 -->
<!-- ```{r Modeling4, message=FALSE, warning=FALSE} -->
<!-- system.time({mod_gam4 = gam(accident  ~ -->
<!--                               s(class1_speed, k = 12) +  -->
<!--                              s(DISTANCE_VIOLATIONS, k = 12) + -->
<!--                              # s(police_enforcment, k = 12) + -->
<!--                              s(JMONTH) + -->
<!--                              s(avg_class2_speed, k = 12) + -->
<!--                              s(avg_class3_speed, k = 12) + -->
<!--                              # ti(class4, TAG, bs = 'fs') + -->
<!--                              ti(class2_speed, class2, k = 12) + -->
<!--                              ti(class3_speed, class3, k = 12) + -->
<!--                              s(DISTANCE_VIOLATIONS, class4, k = 12) + -->
<!--                              # ti(HEAVY_VEHICLES, IS_HOLIDAY, bs = 'fs') + -->
<!--                              s(police_enforcment, IS_HOLIDAY, bs = 'fs') + -->
<!--                              ti(police_enforcment, JMONTH, bs = 'fs') -->
<!--                            ,data = df, select = T, family = 'poisson') -->
<!-- beepr::beep(2) -->
<!-- }) -->
<!-- summary(mod_gam4) -->
<!-- # plot_gam_check(mod_gam4, scatter = TRUE) -->
<!-- ``` -->
<!-- # Modele 5 -->
<!-- ```{r Modelin5, message=FALSE, warning=FALSE} -->
<!-- system.time({mod_gam5 = gam(accident  ~ -->
<!--                               s(class1_speed, k = 12) +  -->
<!--                              s(DISTANCE_VIOLATIONS, k = 12) + -->
<!--                              # s(police_enforcment, k = 12) + -->
<!--                              s(JMONTH) + -->
<!--                              s(avg_class2_speed, k = 12) + -->
<!--                              s(avg_class3_speed, k = 12) + -->
<!--                              # ti(class4, TAG, bs = 'fs') + -->
<!--                              ti(class2_speed, class2, k = 12) + -->
<!--                              ti(class3_speed, class3, k = 12) + -->
<!--                              # s(DISTANCE_VIOLATIONS, class2, k = 12) + -->
<!--                              s(DISTANCE_VIOLATIONS, class4, k = 12) + -->
<!--                              # ti(HEAVY_VEHICLES, IS_HOLIDAY, bs = 'fs') + -->
<!--                              s(police_enforcment, IS_HOLIDAY, bs = 'fs') + -->
<!--                              ti(police_enforcment, JMONTH, bs = 'fs') -->
<!--                            ,data = df, select = T, family = 'poisson') -->
<!-- beepr::beep(5) -->
<!-- }) -->
<!-- summary(mod_gam5) -->
<!-- # plot_gam_check(mod_gam5, scatter = TRUE) -->
<!-- ``` -->


```{r plot.1}
# AIC(mod_gam, mod_gam2, mod_gam3, mod_gam4, mod_gam5)
plot_gam_3d(
   model      = mod_gam,
   main_var   = SPEED_VIOLATIONS,
   # main_var   = SPEED_VIOLATIONS,
   second_var = DISTANCE_VIOLATIONS,
   palette    = 'davos',
   direction  = 1
)
```


### Model checking
```{r gam_check, fig.show='hide'}
test = capture.output(gam.check(mod_gam, k.rep = 1000))
cat(paste0(test, collapse = '\n'))
```
```{r gam_check2, fig.asp=.75}
plot_gam_check(mod_gam, scatter = TRUE)
```



<!-- ### Concurvity -->

<!-- ```{r concurvity} -->
<!-- knitr::kable(concurvity(mod_gam) %>% -->
<!--   as_tibble(rownames = 'type')) -->
<!-- ``` -->